---
title: "Exploratory Data Analysis — Presidential Polling and Results"
subtitle: "Corner Team"
author: "Isobel Cusack, Caitlin Knowles, Linh Chi Nguyen, Manjun Xu"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    embed-resources: false
    theme: cosmo

execute:
  warning: false
  message: false
  echo: true
---

# Exploratory Data Analysis - Presidential Polling and Results

## **Introduction**

This report supplies an exploratory data analysis that aims to clean and explore data surrounding pre-election polling and historical election results of the United States. Trends in pre-1electoral polling combined with voting results will be uncovered using data summaries and visualisations with the goal of identifying plausible research questions that could be further explored. The scope of the report is to explore historical election results at a state level to identify shifts in major party support over time and understand pre-election polling characteristics and how they impact the accuracy of predicting election results.

### **Overview of Datasets**

The following analysis will be performed using three datasets containing presidential polling and election result data:

The first two datasets (State-Level Pre-Election Polls (2012) and State-Level Pre-Election Polls (2016)) contain pre-election polling data from the 2012 (Obama/Romney) and 2016 (Trump/Clinton) US elections. The dataset contains over 1000 entries each and 14 and 16 variables, respectively. Some of the variables include method of polling, sample size of the poll, support for each candidate and number of undecided voters. The datasets contain aggregated polling data from a range of sources collated under the Huffington Post Pollster (pollstR package no longer available) as a part of the rOpenGov project. The data was collected under the Creative Commons Attribution-NonCommercial-ShareAlike 3.0 Unported License, and the pollstR package was released under the GPL-2. These licenses require the user to give appropriate credit, not commercialise the material and distribute any contributions under the same licence as the original with no additional restrictions.

The other dataset (Certified Presidential Returns 1976 - 2024) contains almost 50 years of state-level election results. The dataset contains 4822 rows and 15 columns. It includes variables such as the state and year, total state votes, party and candidate. This dataset was obtained from the MIT Election Data and Science Lab (MEDSL) through the Harvard Dataverse. The data was released in 2017 under the public domain dedication (CCO 1.0), which has no requirements other than scholarly citation (see references).

The three raw datasets have been included in the GitHub repository for reproducibility purposes.

### Limitations

While completing this report, the following limitations must be considered:

The code book for the 1976-2024 election results data is only updated until the year 2020. This means that changes in data collection, recording and formatting procedures are not recorded for the most recent election results. Such issues can impact reproducibility as there is no proper documentation of the processes used.

All three of the datasets contain missing data, which can limit the use of the datasets, particularly when the missingness or errors cannot be resolved. The polling datasets contain structural missing values, most prominently in third-party votes. There is also missingness and inconsistent formatting in the election results dataset, particularly in earlier elections. Not all of these issues can be resolved through data cleaning, meaning the exploration of data may not be an entirely accurate representation of polling and electoral result statistics.

The pollstR package has been archived since 2020, and the HuffPost Pollster API no longer exists, amplifying the need to keep the raw datasets in the repository, as the data can no longer be accessed or reacquired through the original source. This guarantees the reproducibility of the code.

## Data Cleaning

## Exploratory Data Analysis

```{r}
# Loading packages
library(tidyverse)
library(lubridate)
library(skimr)

# Loading datasets
pres_df <- read_csv("data/raw/1976-2024-president.csv")
polls_combined_clean <- read.csv("data/clean/state_polls_combined_clean.csv")
polls_2012_clean <- read_csv("data/clean/state_polls_2012_clean.csv")
polls_2016_clean <- read.csv("data/clean/state_polls_2016_clean.csv")

```

```{r}
# Cleaning president data (1976-2024-president)
pres_clean <- pres_df %>%
  select(-notes) # removes the column "notes" as every row conatins NA 

party_trends <- pres_df %>%
# Replace missing values in party_simplified
  mutate(party_simplified = ifelse(is.na(party_simplified), "OTHER", party_simplified)) %>%
# Calculate national total votes per year and candidate votes per party per year
  group_by(year, party_simplified) %>%
  summarise(party_votes = sum(candidatevotes, na.rm = TRUE), .groups = "drop_last") %>%
  mutate(
    total_year_votes = sum(party_votes),
    national_pct = (party_votes / total_year_votes) * 100
  ) %>%
  ungroup() %>%
 # Lump parties into Democrat, Republican, and Other
  mutate(
    party_simplified = case_when(
      party_simplified %in% c("DEMOCRAT", "REPUBLICAN") ~ party_simplified,
      TRUE ~ "OTHER"
    )
  ) %>%
# Aggregate again: Combine Libertarian % and Other % into a single "other" value per year
  group_by(year, party_simplified) %>%
  summarise(national_pct = sum(national_pct), .groups = "drop")

# Create the line plot
ggplot(party_trends, aes(x = year, y = national_pct, color = party_simplified, group = party_simplified)) +
  geom_line(linewidth = 1) + 
  geom_point(size = 2.5) + 
  scale_color_manual(
    values = c(
      "DEMOCRAT" = "#2E5B88", 
      "REPUBLICAN" = "#D9534F", 
      "OTHER" = "#777777"
    )
# red and blue chosen as they usually correspod to the two major parties
  ) +
  scale_x_continuous(breaks = seq(1976, 2024, by = 8)) + 
  labs( # Adding title and axis labels
    title = "U.S. Presidential Election Vote Share (1976–2024)",
    x = "Election Year",                                      
    y = "National Vote Percentage (%)",                       
    color = "Party"                                              ) +
  theme_minimal()
```

Line plot describing the support of each of the major parties from 1974 to 2024 US elections along with the less prominent parties (such as libertarian) grouped together. The percentage of support changes year to year, showing no significant pattern. It can be observed that the margin of percentage support is reduced in more recent years.

```{r}
# creating a density plot comparing proportion of undecided voters in 2012 vs 2016
polls_combined_clean %>%
  mutate(
    Year = case_when( # changing the year column to also include the two major candidates
      year == 2012 ~ "2012 (Obama/Romney)",
      year == 2016 ~ "2016 (Clinton/Trump)"
    )
  ) %>%ggplot(aes(x = Undecided, fill = Year) # Undecided % on X-axis, color fill grouped by election cycle
  ) +
  geom_density(alpha = 0.5) + # Create smoothed density distribution curves
  scale_fill_manual(
    values = c( # using the same colours to reprent the major parties as done previously
      "2012 (Obama/Romney)" = "#2B5C8F", 
      "2016 (Clinton/Trump)" = "#E74C3C"
    )
  ) +
  labs( # adding title and axis labels
    title = "Distribution of Undecided Voters: 2012 vs. 2016 Polls", # Add title
    x = "Undecided Percentage (%)",                                  # Label X-axis
    y = "Density",                                                   # Label Y-axis
    fill = "Election Cycle"                                          # Legend header title
  ) +
  theme_minimal()
```

Density plot showing the percentage of undecided voters in the 2012 and 2016 elections. There is an overall significantly lower percentage of undecided voters in the 2012 polls in comparison to the 2016 polls.

```{r}
polls_combined_clean %>%
  mutate(
    Year = case_when( # changing the year column to also include the two major candidates
      year == 2012 ~ "2012 (Obama/Romney)",
      year == 2016 ~ "2016 (Clinton/Trump)"
    )
  ) %>%
  filter(!is.na(days_to_election)) %>% # Exlcuding NA inputs
  filter(days_to_election <= 250) %>% # Only showing days 250 or less until the election date
  # Creating density plot for density of polls conducted in relation to the number of dat left until the election date
  ggplot(aes(x = days_to_election, fill = Year)) +
  geom_density(alpha = 0.5) +
  scale_x_reverse() +
  scale_fill_manual(
    values = c(
      "2012 (Obama/Romney)" = "#2B5C8F", 
      "2016 (Clinton/Trump)" = "#E74C3C"
    )
  ) +
  labs( # Adding title and axis labels
    title = "Density of Polling Frequency over Time",
    x = "Days Remaining to Election Day",
    y = "Density of Polling Observations",
    fill = "Election Cycle"
  ) +
  theme_minimal()

```

Density plot describing the density of polling data collected in relation to the remaining days until the 2012 and 2016 elections. Both years follow a similar trajectory of low amounts of polling at dates further from the election, then a spike as election day approaches. The spike in polling in 2016 is more significant than in 2012.

## Research Questions

## References

*Deed - Attribution-NonCommercial-ShareAlike 3.0 Unported - Creative Commons*. (2026). Creativecommons.Org. https://creativecommons.org/licenses/by-nc-sa/3.0/deed.en

*GNU General Public License version 2*. (2006, October 31). Open Source Initiative. https://opensource.org/license/gpl-2.0

MIT Election Data and Science Lab. 2017. “U.S. President 1976–2024.” Harvard Dataverse. [**https://doi.org/10.7910/DVN/42MVDXLinks to an external site.**](https://doi.org/10.7910/DVN/42MVDX). Data retrieved: 27th July 2026. 
