---
title: "Exploratory Data Analysis — Presidential Polling and Results"
subtitle: "Corner Team"
author: "Isobel Cusack, "
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    embed-resources: false
    theme: cosmo

execute:
  warning: false
  message: false
  echo: true
---
*1. Introduction*
```{r}
#| echo: false
library(tidyverse)
library(here)

polls_clean <- read_csv(
  here("data", "clean", "state_polls_combined_clean.csv"),
  show_col_types = FALSE
)

polls_long <- read_csv(
  here("data", "clean", "state_polls_combined_long.csv"),
  show_col_types = FALSE
)
```
**2. Description of Data Cleaning**
*2.1 Poll data cleaning*

The 2012 and 2016 polling datasets were cleaned using a fully reproducible R script. Election year and state were extracted from poll_info, state naming was standardised across years, and polling dates were converted to Date variables. We also created days_to_election to represent the timing of each poll relative to Election Day. poll_id was treated as an identifier rather than a quantitative variable, and a unique poll_key was created using year, poll ID and state.

Invalid sentinel values in sample_size (-1) were recoded as missing. The Other response column contained the text "Not included in poll" as well as numeric percentages; these entries were recoded to NA before converting the variable to numeric.

*2.2 Missingness and structural missingness*

Missing values were not automatically replaced with zero because their meanings differed across variables. For example, missing values for Other, Johnson and McMullin often indicate that the response category or candidate was not included in that poll, rather than zero support. Missing partisan_affiliation for nonpartisan polls was treated as not applicable and recorded as "None".

*2.3 Combining 2012 and 2016 data*

Candidate columns differed between 2012 and 2016, so each dataset was also reshaped into long format with one row per state-poll-candidate. Candidate names were additionally grouped into common categories such as Democrat, Republican, Undecided and Other. The cleaned datasets were then combined into both wide and long versions.

*2.4 Cleaning limitations*

Some structural missingness remains because not every poll reported the same candidates or response categories. A small number of polls also had reported response percentages summing to more than 100%; these were flagged for review rather than automatically altered. In addition, state coverage, pollster representation and polling frequency are uneven, so the combined polling data should not be interpreted as a random sample of U.S. states or voters.

**3. Exploratory Data Analysis**
*3.1 Descriptive overview: Summary statistics table*

```{r}
#| echo: false
# Summary of poll-level data
poll_summary <- polls_clean |>
  group_by(year) |>
  summarise(
    `State-poll records` = n(),
    `State/DC units covered` = n_distinct(state),
    `Pollsters` = n_distinct(pollster),
    `Median sample size` = median(sample_size, na.rm = TRUE),
    `Median days before election` = median(days_to_election, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = -year,
    names_to = "Measure",
    values_to = "value"
  ) |>
  pivot_wider(
    names_from = year,
    values_from = value
  )
show(poll_summary)
```

```{r}
#| echo: false
# Median undecided share
undecided_summary <- polls_long |>
  filter(candidate == "Undecided") |>
  group_by(year) |>
  summarise(
    value = median(pct, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    Measure = "Median reported undecided share"
  ) |>
  pivot_wider(
    names_from = year,
    values_from = value
  )
```

```{r}
#| echo: false
# Combine
summary_table <- bind_rows(
  poll_summary,
  undecided_summary
) |>
  rename(
    `2012 polls` = `2012`,
    `2016 polls` = `2016`
  )

summary_table <- summary_table |>
  mutate(
    `2012 polls` = if_else(
      Measure == "Median reported undecided share",
      paste0(`2012 polls`, "%"),
      as.character(`2012 polls`)
    ),
    `2016 polls` = if_else(
      Measure == "Median reported undecided share",
      paste0(`2016 polls`, "%"),
      as.character(`2016 polls`)
    )
  )

summary_table
```

The 2016 dataset contains more state-poll records, covers all 51 state/DC units, and includes more pollsters than the 2012 dataset. Median poll sample sizes are similar across the two elections. However, the typical 2016 poll was conducted closer to Election Day, and the median reported undecided share was higher in 2016.

*3.4 Accuracy and proximity to Election Day*

```{r}
#| echo: false
results_raw <- read_csv(
  here("data", "raw", "1976-2024-president.csv"),
  show_col_types = FALSE
)
```

```{r}
#| echo: false
# Create actual election margin
actual_results <- results_raw |>
  filter(
    year %in% c(2012, 2016),
    party_simplified %in% c("DEMOCRAT", "REPUBLICAN")
  ) |>
  group_by(year, state, party_simplified) |>
  summarise(
    votes = sum(candidatevotes),
    totalvotes = first(totalvotes),
    .groups = "drop"
  ) |>
  mutate(vote_share = votes / totalvotes * 100) |>
  select(year, state, party_simplified, vote_share) |>
  pivot_wider(
    names_from = party_simplified,
    values_from = vote_share
  ) |>
  mutate(
    actual_margin = DEMOCRAT - REPUBLICAN,
    state = str_to_lower(state),
    state = str_replace_all(state, " ", "-"),
    state = if_else(
      state == "district-of-columbia",
      "washington-dc",
      state
    )
  ) |>
  select(year, state, actual_margin)
```

```{r}
#| echo: false
# Create polling error
poll_accuracy <- polls_clean |>
  mutate(
    poll_margin = case_when(
      year == 2012 ~ Obama - Romney,
      year == 2016 ~ Clinton - Trump
    )
  ) |>
  left_join(actual_results, by = c("year", "state")) |>
  mutate(
    absolute_error = abs(poll_margin - actual_margin)
  )
```

```{r}
#| echo: false
poll_accuracy |>
  ggplot(aes(
    x = days_to_election,
    y = absolute_error
  )) +
  geom_point(
    alpha = 0.12,
    size = 1.3
  ) +
  geom_smooth(
    se = FALSE,
    linewidth = 1
  ) +
  scale_x_reverse(
    breaks = c(600, 500, 400, 300, 200, 100, 0)
  ) +
  facet_wrap(
    ~ year,
    labeller = labeller(
      year = c(
        `2012` = "2012",
        `2016` = "2016"
      )
    )
  ) +
  labs(
    title = "Polling accuracy improves as Election Day approaches",
    subtitle = "Absolute state-level polling error in the 2012 and 2016 U.S. presidential elections",
    x = "Days before Election Day",
    y = "Absolute polling error (percentage points)",
    caption = "Lower values indicate greater polling accuracy."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
  )
```
Figure 3. Absolute state-level polling error by days before Election Day in the 2012 and 2016 presidential elections. Smaller values indicate greater polling accuracy.
