---
title: "Exploratory Data Analysis — Presidential Polling and Results"
subtitle: "Corner Team"
author: "Isobel Cusack, "
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    embed-resources: false
    theme: cosmo

execute:
  warning: false
  message: false
  echo: true
---
*Introduction*

```{r}
# Loading packages
library(tidyverse)
library(lubridate)
library(skimr)

# Loading datasets
pres_df    <- read_csv("data/raw/1976-2024-president.csv")
polls_2012 <- read_csv("data/raw/state_polls_2012.csv")
polls_2016 <- read_csv("data/raw/state_polls_2016.csv")

pres_df

```

```{r}
# Cleaning president data (1976-2024-president)
pres_clean <- pres_df %>%
  select(-notes) # removes the column "notes" as every row conatins NA 

skim(pres_clean)
```

```{r}
party_trends <- pres_clean %>% 
  group_by(year, party_simplified) %>% # Group data by election year and simplified party
  summarise(
    total_party_votes = sum(candidatevotes), # Sum raw votes for each party per year
    .groups = "drop" # Ungroup data after calculating summary
  ) %>%
  group_by(year) %>% # Regroup by year to calculate national percentages
  mutate(
    national_pct = (total_party_votes / sum(total_party_votes)) * 100 # Divide party total by overall year total
  )
party_trends
```

```{r}
# Plot 1: Line chart showing vote percentage trends by party (1976-2024)
p1 <- ggplot(
    party_trends, 
    aes(x = year, y = national_pct, color = party_simplified) # Set year as X, vote % as Y, color by party
  ) +
  geom_line(linewidth = 1) + # Draw trend lines with a line width of 1.2
  geom_point(size = 2.5) + # Add points at each election year with a point size of 2.5
  scale_color_manual(
    # Custom color palette matching party branding
    values = c(
      "DEMOCRAT" = "#2E5B88", 
      "REPUBLICAN" = "#D9534F", 
      "LIBERTARIAN" = "#F0AD4E", 
      "OTHER" = "#777777"
    )
  ) +
  scale_x_continuous(breaks = seq(1976, 2024, by = 8)) + # Set X-axis ticks every 8 years (e.g., 1976, 1984, ...)
  labs(
    title = "U.S. Presidential Election Vote Share (1976–2024)", # Add chart main title and label axis and legend
    x = "Election Year",                                         
    y = "National Vote Percentage (%)",                          
    color = "Party"                                              
  ) +
  theme_minimal() # Apply clean background theme with minimal grid lines
print(p1)
```

```{r}
# Clean and prepare 2012 polling data
polls_2012_clean <- polls_2012 %>% 
  filter(sample_size > 0) %>% # Filter out bad data entries (removes sample_size = -1)
  mutate(
    start_date = mdy(start_date), 
    end_date   = mdy(end_date), # Convert dates string to date object
    obama_margin = Obama - Romney, # Calculate point margin (positive = Obama leading, negative = Romney leading)
    
# Extract the state name from the 'poll_info' string and creating a seperate column for "state"
    state = str_extract(poll_info, "(?<=2012-)[a-z-]+(?=-president)") %>% 
            str_replace_all("-", " ") # Replace hyphens with spaces 
  )
polls_2012_clean
```

```{r}
# Display summary statistics for key numerical variables
polls_2012_clean %>%
  select(Obama, Romney, Undecided, obama_margin, sample_size) %>% # Keep relevant columns
  summary()
```

```{r}
p2 <- ggplot(
    polls_2012_clean, 
    aes(x = mode, y = obama_margin, fill = mode) # Group X by mode, Y as margin, color fill by mode
  ) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) + # Create boxplot with 70% opacity and hide  legend
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") + # Add reference line at Y = 0 (tied race)
  labs( # Add title and label axis
    title = "2012 Poll Margin (Obama - Romney) by Polling Mode", 
    x = "Polling Mode",                                          
    y = "Obama Margin (% Points)"                                
  ) +
  theme_minimal() +
  coord_flip()      # Horizontal flip so polling modes display clearly on vertical axis

print(p2) 
```

```{r}
# Clean and prepare 2016 polling data
polls_2016_clean <- polls_2016 %>% 
  filter(sample_size > 0) %>% # Filter out bad data entries 
  mutate(
    start_date = mdy(start_date), 
    end_date   = mdy(end_date),   # Convert text date to Date class
    
    clinton_margin = Clinton - Trump, # Calculate point margin (positive = Clinton lead, negative = Trump lead)
    
    # Extract state name from poll_info string and create "state" column
    state = str_extract(poll_info, "(?<=2016-)[a-z-]+(?=-president)") %>% 
            str_replace_all("-", " ")                             # Convert dashes into spaces
  )
polls_2016_clean
```

```{r}
polls_2016_clean %>%
  select(Clinton, Trump, Undecided, clinton_margin, Johnson, sample_size) %>% # Select key numeric columns
  summary()
```


```{r}
comp_df <- bind_rows(
  polls_2012_clean %>% select(Undecided) %>% mutate(Year = "2012 (Obama/Romney)"), # Extract 2012 Undecided and tag cycle year
  polls_2016_clean %>% select(Undecided) %>% mutate(Year = "2016 (Clinton/Trump)")  # Extract 2016 Undecided and tag cycle year
) %>% 
  filter(!is.na(Undecided)) # Exclude missing values
comp_df
```

```{r}
# Plot 4: Density plot comparing proportion of undecided voters in 2012 vs 2016
p4 <- ggplot(
    comp_df, 
    aes(x = Undecided, fill = Year) # Undecided % on X-axis, color fill grouped by election cycle
  ) +
  geom_density(alpha = 0.5) + # Create smoothed density distribution curves with 50% opacity
  scale_fill_manual(
    values = c(
      "2012 (Obama/Romney)" = "#2B5C8F", 
      "2016 (Clinton/Trump)" = "#E74C3C"
    )
  ) +
  labs(
    title = "Distribution of Undecided Voters: 2012 vs. 2016 Polls", # Add title
    x = "Undecided Percentage (%)",                                  # Label X-axis
    y = "Density",                                                   # Label Y-axis
    fill = "Election Cycle"                                          # Legend header title
  ) +
  theme_minimal()

print(p4) # Display plot p4
```

p2 <- ggplot(
  mode_accuracy,
  aes(
    x = mean_abs_error,
    y = reorder(mode, mean_abs_error)
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = round(mean_abs_error, 2)),
    hjust = -0.2,
    size = 3.5
  ) +
  labs(
    title = "Polling Accuracy by Survey Mode",
    subtitle = "Lower error indicates estimates closer to actual election results",
    x = "Mean Absolute Polling Error (percentage points)",
    y = "Survey Mode"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(
      size = 15,
      face = "bold"
    ),
    plot.subtitle = element_text(
      size = 10
    ),
    panel.grid.major.y = element_blank()
  ) +
  expand_limits(
    x = max(mode_accuracy$mean_abs_error) * 1.12
  )

print(p2)




