---
title: "Exploratory Data Analysis — Presidential Polling and Results"
subtitle: "Corner Team"
author: "Isobel Cusack, "
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    embed-resources: false
    theme: cosmo

execute:
  warning: false
  message: false
  echo: true
---

# Exploratory Data Analysis - Presedential Polling and Results

## **Introduction**

This report supplies an exporatory data analysis that aims to clean and explore data surrounding pre-election polling and historical election results of the United States. Trends in pre electoral polling combined with voting results will be uncovered using data summaries and visualisations with the goal of identifying plausible research questions that could be further explored. The scope of the report is to explore historcal election results at a state level to identify shifts in major party support over time and understand pre-election polling characteristics and how they impact the accuracy of predicting election results.

### **Overveiw of Datasets**

The following analysis will be perfromed using three datasets containing presidential polling and election result data:

The first two datasets (State-Level Pre-Election Polls (2012) and State-Level Pre-Election Polls (2016)) contain pre-election polling data from the 2012 (Obama/Romney) and 2016 (Trump/Clinton) US elections. The dataset contains over 1000 entries each and 14 and 16 varaibles respectively. Some of the variables include, method of polling, sample size of the poll, support for each candidate and number of undecided voters. The datasets contains agregated polling data from a range of sources collated under the Huffington Post Pollster (pollstR package no longer available) as a part of rOpenGov project. The data was collected under the Creative Commons Attribution-NonCommercial-ShareAlike 3.0 Unported License and the pollstR package was released uder the GPL-2. These licenses require the user to give appropraite credit, not commercialise the material and distribute any contributions under the same lisence as the original with no additional restrictions.

The other data set (Certified Presidential Returns 1976 - 2024) conatins almost 50 years of state level election results. The dataset contains 4822 rows and 15 columns. it includes variables such as the state and year, total state votes, party and candidate. This dataset was obtained from the MIT Election Data and Science Lab (MEDSL) through the Harvard Dataverse. The data was released in 2017 under the public domain dedication (CCO 1.0) which has no requirements other than scholarly citation (see references).

The three raw datasets have been included in the Github repository for reproducibility purposes.

### Limitations

While completing this report, the following limitations must be considered:

The code book for the 1976-2024 election results data is only updated until the year 2020. This means that changes in data collection, recording and formatting proceedures are not recorded for the most recent election results. Such issues can impact the reproducibility as there is no propper documentation of the proccesses used.

All three of the datasets conatin missing data which can limit the use of the datasets particularly is the missingness or errors cannot be resolved. the polling datasets contain sturctural missing values most prominently in third party votes. There is also missingness and inconsistent formatting in the election results dataset particularly in earlier election. Not all of these issues can be resolved through data cleaning meaning the exploration of data may not be an entirely accurate representation of polling and electoral result statistics.

The pollstR package has been archived since 2020 and the Huffpost Pollster API no longer exists, amplifying the need to keep the raw datasets in the repository as the data can no longer be accessed or reaquired through the original source. This garauntees the reproducibility of the code.

## Data Cleaning

```{r}
# Loading packages
library(tidyverse)
library(lubridate)
library(skimr)

# Loading datasets
pres_df    <- read_csv("data/raw/1976-2024-president.csv")
polls_2012 <- read_csv("data/raw/state_polls_2012.csv")
polls_2016 <- read_csv("data/raw/state_polls_2016.csv")

polls_2016

```

```{r}
# load in clean datasets

polls_combined_clean <- read.csv("data/clean/state_polls_combined_clean.csv")

polls_2012_clean <- read_csv("data/clean/state_polls_2012_clean.csv")
polls_2016_clean <- read.csv("data/clean/state_polls_2016_clean.csv")

polls_combined_clean

```

```{r}
# Cleaning president data (1976-2024-president)
pres_clean <- pres_df %>%
  select(-notes) # removes the column "notes" as every row conatins NA 

skim(pres_clean)
```

```{r}
party_trends <- pres_clean %>% 
  group_by(year, party_simplified) %>% # Group data by election year and simplified party
  summarise(
    total_party_votes = sum(candidatevotes), # Sum raw votes for each party per year
    .groups = "drop" # Ungroup data after calculating summary
  ) %>%
  group_by(year) %>% # Regroup by year to calculate national percentages
  mutate(
    national_pct = (total_party_votes / sum(total_party_votes)) * 100 # Divide party total by overall year total
  )
party_trends
```

```{r}
party_trends <- pres_df %>%
  # Replace missing values in party_simplified
  mutate(party_simplified = ifelse(is.na(party_simplified), "OTHER", party_simplified)) %>%
  # Calculate national total votes per year and candidate votes per party per year
  group_by(year, party_simplified) %>%
  summarise(party_votes = sum(candidatevotes, na.rm = TRUE), .groups = "drop_last") %>%
  mutate(
    total_year_votes = sum(party_votes),
    national_pct = (party_votes / total_year_votes) * 100
  ) %>%
  ungroup() %>%
  # Lump parties into DEMOCRAT, REPUBLICAN, and OTHER
  mutate(
    party_simplified = case_when(
      party_simplified %in% c("DEMOCRAT", "REPUBLICAN") ~ party_simplified,
      TRUE ~ "OTHER"
    )
  ) %>%
  # AGGREGATE AGAIN: Combine Libertarian % and Other % into a single OTHER value per year
  group_by(year, party_simplified) %>%
  summarise(national_pct = sum(national_pct), .groups = "drop")

# 3. Create the plot
p1 <- ggplot(party_trends, aes(x = year, y = national_pct, color = party_simplified, group = party_simplified)) +
  geom_line(linewidth = 1) + 
  geom_point(size = 2.5) + 
  scale_color_manual(
    values = c(
      "DEMOCRAT" = "#2E5B88", 
      "REPUBLICAN" = "#D9534F", 
      "OTHER" = "#777777"
    )
  ) +
  scale_x_continuous(breaks = seq(1976, 2024, by = 8)) + 
  labs(
    title = "U.S. Presidential Election Vote Share (1976–2024)",
    x = "Election Year",                                         
    y = "National Vote Percentage (%)",                           
    color = "Party"                                              
  ) +
  theme_minimal()

print(p1)
```

Figure X: Line plot describing the support of each of the major parties from 1974 to 2024 US elections along with the less prominent parties (such as libertarian) grouped together. The percentage of suppport changes year to year showing no significant pattern. It can be observed that the margin of percentage support is reduced in more recent years.

```{r}
# Clean and prepare 2012 polling data
polls_2012_clean <- polls_2012 %>% 
  filter(sample_size > 0) %>% # Filter out bad data entries (removes sample_size = -1)
  mutate(
    start_date = mdy(start_date), 
    end_date   = mdy(end_date), # Convert dates string to date object
    obama_margin = Obama - Romney, # Calculate point margin (positive = Obama leading, negative = Romney leading)
    
# Extract the state name from the 'poll_info' string and creating a seperate column for "state"
    state = str_extract(poll_info, "(?<=2012-)[a-z-]+(?=-president)") %>% 
            str_replace_all("-", " ") # Replace hyphens with spaces 
  )
polls_2012_clean
```

```{r}
# Display summary statistics for key numerical variables
polls_2012_clean %>%
  select(Obama, Romney, Undecided, obama_margin, sample_size) %>% # Keep relevant columns
  summary()
```

```{r}
p2 <- ggplot(
    polls_2012_clean, 
    aes(x = mode, y = obama_margin, fill = mode) # Group X by mode, Y as margin, color fill by mode
  ) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) + # Create boxplot with 70% opacity and hide  legend
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") + # Add reference line at Y = 0 (tied race)
  labs( # Add title and label axis
    title = "2012 Poll Margin (Obama - Romney) by Polling Mode", 
    x = "Polling Mode",                                          
    y = "Obama Margin (% Points)"                                
  ) +
  theme_minimal() +
  coord_flip()      # Horizontal flip so polling modes display clearly on vertical axis

print(p2) 
```

```{r}
# Clean and prepare 2016 polling data
polls_2016_clean <- polls_2016 %>% 
  filter(sample_size > 0) %>% # Filter out bad data entries 
  mutate(
    start_date = mdy(start_date), 
    end_date   = mdy(end_date),   # Convert text date to Date class
    
    clinton_margin = Clinton - Trump, # Calculate point margin (positive = Clinton lead, negative = Trump lead)
    
    # Extract state name from poll_info string and create "state" column
    state = str_extract(poll_info, "(?<=2016-)[a-z-]+(?=-president)") %>% 
            str_replace_all("-", " ")                             # Convert dashes into spaces
  )
polls_2016_clean
```

```{r}
polls_2016_clean %>%
  select(Clinton, Trump, Undecided, clinton_margin, Johnson, sample_size) %>% # Select key numeric columns
  summary()
```

```{r}
comp_df <- bind_rows(
  polls_2012_clean %>% select(Undecided) %>% mutate(Year = "2012 (Obama/Romney)"), # Extract 2012 Undecided and tag cycle year
  polls_2016_clean %>% select(Undecided) %>% mutate(Year = "2016 (Clinton/Trump)")  # Extract 2016 Undecided and tag cycle year
) %>% 
  filter(!is.na(Undecided)) # Exclude missing values
comp_df
```

```         
```

```{r}
# Plot 4: Density plot comparing proportion of undecided voters in 2012 vs 2016
p4 <- polls_combined_clean %>%
  mutate(
    Year = case_when(
      year == 2012 ~ "2012 (Obama/Romney)",
      year == 2016 ~ "2016 (Clinton/Trump)"
    )
  ) %>%ggplot(aes(x = Undecided, fill = Year) # Undecided % on X-axis, color fill grouped by election cycle
  ) +
  geom_density(alpha = 0.5) + # Create smoothed density distribution curves with 50% opacity
  scale_fill_manual(
    values = c(
      "2012 (Obama/Romney)" = "#2B5C8F", 
      "2016 (Clinton/Trump)" = "#E74C3C"
    )
  ) +
  labs(
    title = "Distribution of Undecided Voters: 2012 vs. 2016 Polls", # Add title
    x = "Undecided Percentage (%)",                                  # Label X-axis
    y = "Density",                                                   # Label Y-axis
    fill = "Election Cycle"                                          # Legend header title
  ) +
  theme_minimal()

print(p4) # Display plot p4
```

Figure X: Density plot showing the percentage of undecided voters in the 2012 and 2016 elections. There is an overall significantly lower percentage of undecided voters in the 2012 polls in comparison to the 2016 polls.

```{r}
p4 <- polls_combined_clean %>%
  mutate(
    Year = case_when(
      year == 2012 ~ "2012 (Obama/Romney)",
      year == 2016 ~ "2016 (Clinton/Trump)"
    )
  ) %>%
  filter(!is.na(days_to_election)) %>%
  filter(days_to_election <= 250) %>%
  ggplot(aes(x = days_to_election, fill = Year)) +
  geom_density(alpha = 0.5) +
  scale_x_reverse() +
  scale_fill_manual(
    values = c(
      "2012 (Obama/Romney)" = "#2B5C8F", 
      "2016 (Clinton/Trump)" = "#E74C3C"
    )
  ) +
  labs(
    title = "Density of Polling Frequency over Time",
    x = "Days Remaining to Election Day",
    y = "Density of Polling Observations",
    fill = "Election Cycle"
  ) +
  theme_minimal()

print(p4)
```

Figure X: Density plot describing the desity of polling data collected in relation to the remaining days until the 2012 and 2016 election. Both years follow a similar trajectory of low amounts of polling at dates further from the election then a spike as election day approaches. The spike in polling in 2016 is more significant that 2012.

## References

*Deed - Attribution-NonCommercial-ShareAlike 3.0 Unported - Creative Commons*. (2026). Creativecommons.Org. https://creativecommons.org/licenses/by-nc-sa/3.0/deed.en

*GNU General Public License version 2*. (2006, October 31). Open Source Initiative. https://opensource.org/license/gpl-2.0

MIT Election Data and Science Lab. 2017. “U.S. President 1976–2024.” Harvard Dataverse. [**https://doi.org/10.7910/DVN/42MVDXLinks to an external site.**](https://doi.org/10.7910/DVN/42MVDX). Data retrieved: 27th July 2026. 
