---
title: "Exploratory Data Analysis — Presidential Polling and Results"
subtitle: "Group project template"
author: "Your names here"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    embed-resources: false
    theme: cosmo
# To render a PDF instead, install tinytex ('quarto install tinytex' in the terminal) and change:
# format: pdf
execute:
  warning: false
  message: false
  echo: true
---

# Overview

The purpose of this quarto file is to explore the polling (and eventually results) datasets. 
You should aim to explore (and correct)

1. Problems with the _way_ the dataset is stored that requires wrangling (cleaning, manipulations)
2. The data itself; statistical properties and potential issues (e.g., biases) that it may contain.

Some code snippets are provided but you should add as much more exploratory data analysis and cleaning during this forum and indeed for the EDA Report.

# Setup

```{r}
#| label: setup
# install.packages(c("tidyverse", "here", "lubridate", "glue"))
library(tidyverse)   # dplyr, ggplot2, tidyr, readr, stringr, forcats
library(here)        # project-relative paths; don't use setwd()
library(lubridate)   # dates
library(glue)        # easy string manipulation

raw_dir   <- here("data", "raw")
clean_dir <- here("data", "clean")

# Election day = first Tuesday after the first Monday in November.
election_day <- c("2012" = as.Date("2012-11-06"),
                  "2016" = as.Date("2016-11-08"))
```

`here()` anchors paths at the project root (determined by the .Rproj file), so this renders on any machine
without editing file paths. Keep this `.qmd` at the project root (next to the
`data/` folder).

---

# Source — where did this data come from?

::: {.callout-note}
## Discuss
Before interrogating the data directly, read the data information page and provided codebook.
For **each** of the `.csv` three files, record: who produced it and from
what underlying source; what licence governs it and what that obliges you to do;
and whether it is raw or already processed, and by whom.

Also make notes of any irregularities or nuances in the data you may need to be careful with during analysis.
:::



::: {.callout-note}
## Discuss
The codebook is titled 1976–2020 but the data file runs to 2024. What are the
implications of documentation lagging the data?
:::

*Your answer:*

---

# Load the raw data

```{r}
#| label: load
#| message: false
polls_2016_raw <- read_csv(glue("{raw_dir}/state_polls_2016.csv"))

glimpse(polls_2016_raw)

```

::: {.callout-note}
## Discuss
Did any column read as an unexpected format?
What does that tell you about the contents of the column?
Are there any columns that should be cleaned further? If so, what needs to be done?
:::

*Your answer:*

---

# Who / what is represented?

::: {.callout-note}
## Discuss
State the unit of observation for **each** file in one sentence. (What does a
single row correspond to?)
:::

*Your answer:*

```{r}
#| label: poll-coverage-2016
polls_2016_raw |> summarise(
  n_polls     = n(),
  n_pollster  = n_distinct(pollster),
  first_field = min(start_date, na.rm = TRUE),
  last_field  = max(end_date,   na.rm = TRUE)
)
```

```{r}
#| label: poll-coverage-2012-todo
#| eval: false
# Your turn: produce the same coverage summary for the 2012 polls.
```

```{r}
#| label: election-result-summary
#| eval: false
# Your turn: produce the same coverage summary for the 2012 polls.
```

::: {.callout-note}
## Discuss
What population is captured in a **poll** row versus a **results** row? Who may be *excluded*
from each? For the polling datasets, distinguish the target population (who should be included in the survey) from the sampling frame (who is actually surveyed).
:::

*Your answer:*

---

# How were the variables measured?

::: {.callout-note}
## Discuss
The candidate columns in the polls are reported **percentages, not counts**.
Who chose the denominator? Are they percentages of all respondents, or of
decided voters only? How would you tell?
:::

*Your answer:*

# How much variability is there in variables measured?

::: {.callout-note}
## Discuss
The columns `sample_subpopulation`, `mode`, `partisanship`, and `partisan_affiliation` describe characteristics of the polls. How much variation is in these columns? Are there any potential problems for statistical inference you should note?
:::

*Your answer:*

```{r}
#| label: measurement-meta
polls_2016_raw |> count(sample_subpopulation, sort = TRUE)
```

---

# What procedures were followed during measurement?

::: {.callout-note}
## Discuss
Polls are weighted and adjusted by each pollster using (mostly) undisclosed procedures.
What does the polling dataset tell you about the polling procedures? What decisions/rules were likely to be undisclosed?
How might this affect subsequent analysis?
:::

*Your answer:*

---

# Temporal and spatial variation

The time recording in the polling data is arbitary. It is more informative define a variable that counts down to the election date, as follows.

```{r}
#| label: time-to-election
polls_2016 <- polls_2016_raw |>
  mutate(
    start_date = as.Date(start_date),
    end_date   = as.Date(end_date),
    days_to_election = as.integer(election_day["2016"] - end_date)
  )
```

```{r}
#| label: time-plot
#| fig-cap: "Temporal distribution of polling effort, 2016."
polls_2016 |>
  ggplot(aes(days_to_election)) +
  geom_histogram(binwidth = 7) +
  scale_x_reverse() +
  labs(x = "Days to election (binned by week)", y = "Number of polls")
```

```{r}
#| label: space-coverage-todo
#| eval: false
# Your turn: the state is encoded inside `poll_info`
# (e.g. "2016-washington-d-c-president-..."). If you haven't already, extract it into a `state` and `year` columns
# (try stringr::str_extract), then:
# polls_2016 |> count(state, sort = TRUE) |> print(n = Inf)
```

::: {.callout-note}
## Discuss
Is polling effort uniform over time? Which states are heavily polled and which
are barely polled — is that random, or related to competitiveness? What does it
imply for any nationwide statement? Do your conclusions hold for 2012 as well?
:::

*Your answer:*

---

# Cleaning, tidying, and combining

::: {.callout-note}
## Discuss
The 2012 and 2016 poll files have **different candidate columns** (Obama/Romney
vs Trump/Clinton/Johnson/McMullin) and different column orders. To analyse them
together, reshape to a tidy long form: one row per (poll, candidate). Why is
long form easier to combine and model than the wide form supplied?
:::

*Your answer:*

```{r}
#| label: reshape-2016
#| eval: false # change to true
candidate_cols_2016 <- c("Trump", "Clinton", "Johnson",
                         "McMullin", "Other", "Undecided")

polls_2016_long <- polls_2016 |>
  pivot_longer(cols = all_of(candidate_cols_2016),
               names_to = "candidate", values_to = "pct")

glimpse(polls_2016_long)
```

```{r}
#| label: reshape-bind-todo
#| eval: false
# Your turn:
# 1. Build polls_2012_long the same way (candidate cols: Obama, Romney,
#    Other, Undecided).
# 2. Make sure there is a `year` column in each, then bind_rows() into one `polls_long`.
# 3. Decide how to treat candidates present in one year but not the other.
```

---

# Missingness — what is missing, and how is it recorded?

Analyse the missingness in the combined polling dataset.

::: {.callout-note}
## Discuss
- Is missing `Undecided` a true zero (nobody undecided) or an unknown (not
  reported)? How would you check, and why does it matter?
- A negative `sample_size` is impossible — what value is being used as a
  sentinel for "unknown"? What happens to your means/weights if you leave it in?
- What non-numeric entries appear in `Other`, and what will you do with them?
- For each affected variable: is it MCAR, MAR, or MNAR as far as you can tell,
  and will you drop, recode, impute, or model around it? Every fix lives in
  code, not in the raw file.
:::

*Your answer:*

---

# Bias — who/what is over- or under-represented?

```{r}
#| label: bias-pollster
polls_2016_raw |> count(pollster, sort = TRUE) |> head(15)
```


::: {.callout-note}
## Discuss
- **Coverage:** are the polls a representative sample of *states*, or
  concentrated on a few?
- **Pollster concentration:** if a few firms dominate the rows, whose house
  methods are over-weighted, and how might you account for it (grouping,
  weighting, a pollster effect)?
- **Partisanship:** do partisan-sponsored polls differ systematically? Nuisance
  to remove, or variable to study?
- **Selection:** the data is what someone *chose* to collect (which states,
  which weeks). Name one conclusion this selection could distort.
:::

*Your answer:*

---

# Wrangling data

This `.qmd` file can serve as a place to experiment with data cleaning and exploration. You can add/change it in future forums and when working on the EDA Report assessment.

For clarity and reproducibility you should create a master `poll-cleaning.R` script which starts with the raw data files and creates the cleaned, combined versions of the datasets you will perform EDA on. The code in this file serves as a starting point for this master cleaning script. 

The basic structure of the master cleaning script should be as follows:

```{r}
#| label: read-clean-combine-write
#| eval: false

# load packages
library(...)

# set up directories
raw_data_dir <- here(...)
clean_data_dir <- here(...)

# read in raw data
raw_data <- read_csv(raw_data_dir...)

# clean datasets
clean_data <- raw_data |> mutate(...)

# combine datasets as required
combinded_data <- bind_rows(...)
# save datasets
write_csv(clean_data_dir...)
```

All subsequent EDA and analysis should use the cleaned data.
This document serves as a record of the cleaning and EDA you have tried (but not necessarily included in your report).