---
title: "Exploratory Data Analysis — Presidential Polling and Results"
subtitle: "Group project template"
author: "Your names here"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    embed-resources: false
    theme: cosmo
# To render a PDF instead, install tinytex ('quarto install tinytex' in the terminal) and change:
# format: pdf
execute:
  warning: false
  message: false
  echo: true
---

# Overview

The purpose of this quarto file is to explore the polling (and eventually results) datasets. 
You should aim to explore (and correct)

1. Problems with the _way_ the dataset is stored that requires wrangling (cleaning, manipulations)
2. The data itself; statistical properties and potential issues (e.g., biases) that it may contain.

Some code snippets are provided but you should add as much more exploratory data analysis and cleaning during this forum and indeed for the EDA Report.

# Setup

```{r}
#| label: setup
# install.packages(c("tidyverse", "here", "lubridate", "glue"))
library(tidyverse)   # dplyr, ggplot2, tidyr, readr, stringr, forcats
library(here)        # project-relative paths; don't use setwd()
library(lubridate)   # dates
library(glue)        # easy string manipulation

raw_dir   <- here("data", "raw")
clean_dir <- here("data", "clean")
dir.create(clean_dir, recursive = TRUE, showWarnings = FALSE)

# Election day = first Tuesday after the first Monday in November.
election_day <- c("2012" = as.Date("2012-11-06"),
                  "2016" = as.Date("2016-11-08"))
```

`here()` anchors paths at the project root (determined by the .Rproj file), so this renders on any machine
without editing file paths. Keep this `.qmd` at the project root (next to the
`data/` folder).

---

# Source — where did this data come from?

::: {.callout-note}
## Discuss
Before interrogating the data directly, read the data information page and provided codebook.
For **each** of the `.csv` three files, record: who produced it and from
what underlying source; what licence governs it and what that obliges you to do;
and whether it is raw or already processed, and by whom.
1. state_polls_2012.csv and state_polls_2016.csv

- Originally from HuffPost Pollster, which aggregated US public opinion polls.
- Accessed through the R package pollstR.
- These are processed/compiled polling data, not raw individual survey responses.
- Polling data licence: CC BY-NC-SA 3.0 -> we need to give attribution and follow the non-commercial/share-alike conditions.
- pollstR itself uses GPL-2.0 — this applies to the software, not the polling data.

Also make notes of any irregularities or nuances in the data you may need to be careful with during analysis.
- different pollsters used different methods, samples and weighting procedures.

2. 1976-2024-president.csv
- Contains state-level official presidential election results and follows the MIT Election Data and Science Lab dataset structure. MIT describes its presidential dataset as state-level election returns.
- It is already processed and standardised
- The older 1976–2020 release is associated with the MIT Election Data and Science Lab dataset; because our file now includes 2024, we should confirm the current licence/version before final submission rather than assume the old documentation fully applies.
:::

::: {.callout-note}
## Discuss
The codebook is titled 1976–2020 but the data file runs to 2024. What are the
implications of documentation lagging the data?
:::

- The documentation is older than the dataset.
- Some 2024 values, categories or coding rules may not be documented.
- We should check 2024 separately before assuming the old codebook applies.
- Mention this as a documentation/reproducibility limitation in the report.

---

# Load the raw data

```{r}
#| label: load
#| message: false
polls_2016_raw <- read_csv(glue("{raw_dir}/state_polls_2016.csv"))
polls_2012_raw <- read_csv(glue("{raw_dir}/state_polls_2012.csv"))
result_1976_2024_raw <- read_csv(glue("{raw_dir}/1976-2024-president.csv"))

glimpse(polls_2016_raw)
glimpse(polls_2012_raw)
```

::: {.callout-note}
## Discuss
Did any column read as an unexpected format?
What does that tell you about the contents of the column?
Are there any columns that should be cleaned further? If so, what needs to be done?
:::

- "Other" column is the main problem: it contains numbers and text such as "Not included in poll", so it is read as text.
-> Convert valid values in "Other" to numeric and change "Not included in poll" to NA.
- sample_size contains -1, which is impossible as a real sample size → treat it as missing.
- Make sure start_date and end_date are proper Date variables.
- Extract year and state from poll_info.
- Do not treat repeated poll_ids automatically as duplicate rows, one poll ID can contain results for several states.
```{r}
# Repeated poll_id values are not automatically duplicates because the same poll
# can report results for more than one state. Check exact duplicate rows instead.
tibble(
  dataset = c("2012", "2016"),
  exact_duplicate_rows = c(
    sum(duplicated(polls_2012_raw)),
    sum(duplicated(polls_2016_raw))
  ),
  unique_poll_ids = c(
    n_distinct(polls_2012_raw$poll_id),
    n_distinct(polls_2016_raw$poll_id)
  )
)
```

---

# Who / what is represented?
1. Poll rows
- Target population can be Likely Voters, Registered Voters, or Adults.
- Actual respondents depend on each pollster's sampling frame and survey method.
- People who cannot be reached by that survey method or do not respond may be excluded.

2. Results rows
- Represent actual recorded votes, not opinion poll respondents.
- People who were eligible but did not vote are not represented.

::: {.callout-note}
## Discuss
State the unit of observation for **each** file in one sentence. (What does a
single row correspond to?)
:::

2012 polling file & 2016 polling file
- One row = one state-level poll result, for a pollster × field period × sample population.

Election results file
- One row = one candidate/party result in one state in one election year.
- candidatevotes = votes for that candidate.
- totalvotes = total votes in that state/year.


```{r}
#| label: poll-coverage-2016
polls_2016_raw |> summarise(
  n_polls     = n(),
  n_pollster  = n_distinct(pollster),
  first_field = min(start_date, na.rm = TRUE),
  last_field  = max(end_date,   na.rm = TRUE)
)
```

```{r}
#| label: poll-coverage-2012-todo
#| eval: false
# Your turn: produce the same coverage summary for the 2012 polls.
polls_2012_raw |> summarise(
  n_polls     = n(),
  n_pollster  = n_distinct(pollster),
  first_field = min(start_date, na.rm = TRUE),
  last_field  = max(end_date,   na.rm = TRUE)
)
```

```{r}
#| label: election-result-summary
#| eval: false
result_1976_2024_raw |> summarise(
  n_results     = n(),
  n_state  = n_distinct(state),
  first_field = min(year, na.rm = TRUE),
  last_field  = max(year,   na.rm = TRUE)
)
```

::: {.callout-note}
## Discuss
What population is captured in a **poll** row versus a **results** row? Who may be *excluded*
from each? For the polling datasets, distinguish the target population (who should be included in the survey) from the sampling frame (who is actually surveyed).
:::

*Your answer:*

---

# How were the variables measured?

::: {.callout-note}
## Discuss
The candidate columns in the polls are reported **percentages, not counts**.
Who chose the denominator? Are they percentages of all respondents, or of
decided voters only? How would you tell?
:::

- Candidate columns are percentages, not numbers of people.
- The denominator was determined by each pollster.
- The data suggest percentages are usually based on all respondents, because candidate support + undecided + other often adds to around 100%.
- We should check the original poll methodology before assuming every poll uses exactly the same denominator.

# How much variability is there in variables measured?

::: {.callout-note}
## Discuss
The columns `sample_subpopulation`, `mode`, `partisanship`, and `partisan_affiliation` describe characteristics of the polls. How much variation is in these columns? Are there any potential problems for statistical inference you should note?
:::

Sample population
- 2012: mostly Likely Voters — 967/1,135.
- 2016: mostly Likely Voters — 1,088/1,303.
- Very few polls use Adults.
Survey mode
- 2012: mainly Live Phone and Automated Phone.
- 2016: Internet became the most common method, followed by Live Phone.
Partisanship
- Most polls are Nonpartisan: 2012: 823, 2016: 1,087
- Some are pollster- or sponsor-partisan.
Statistical issue
- Different populations, modes and pollsters may produce systematically different results.
- So observations are not necessarily directly comparable or independent.

```{r}
#| label: measurement-meta
polls_2012_raw |> count(sample_subpopulation, sort = TRUE)
polls_2016_raw |> count(sample_subpopulation, sort = TRUE)

```

---

# What procedures were followed during measurement?
- Pollster
- survey dates
- sample size
- target population
- survey mode
- partisanship

::: {.callout-note}
## Discuss
Polls are weighted and adjusted by each pollster using (mostly) undisclosed procedures.
What does the polling dataset tell you about the polling procedures? What decisions/rules were likely to be undisclosed?
How might this affect subsequent analysis?

:::

Undisclosed procedures:
- Exact sampling procedure
- Weighting method
- Likely-voter model
- Question wording/order
- How non-response was adjusted
- How undecided voters were handled

How might this affect subsequent analysis?

- Two polls can produce different results because of their methodology, not only because voter opinion changed.
- Pollster/mode effects may need to be controlled for in later analysis.
---

# Temporal and spatial variation

The time recording in the polling data is arbitrary. It is more informative define a variable that counts down to the election date, as follows.

```{r}
#| label: clean-polls-wide

candidate_cols_2012 <- c("Obama", "Romney", "Undecided", "Other")
candidate_cols_2016 <- c("Trump", "Clinton", "Johnson",
                         "McMullin", "Other", "Undecided")

# Keep one CLEAN WIDE dataset per election.
# This preserves one row per state-poll and is the safest format for
# poll-count / coverage analyses.

polls_2012_clean <- polls_2012_raw |>
  extract(
    poll_info,
    into = c("year", "state", "contest"),
    regex = "^(\\d{4})-(.*?)-(president|presidential-general-election)-.*$",
    remove = FALSE
  ) |>
  mutate(
    year = as.integer(year),
    state = str_replace(state, "^washington-d-c$", "washington-dc"),
    start_date = as.Date(start_date),
    end_date = as.Date(end_date),
    days_to_election = as.integer(election_day["2012"] - end_date),

    # poll_id is an identifier, not a quantity.
    poll_id = as.character(poll_id),

    # -1 is a sentinel for an unknown sample size.
    sample_size = as.integer(na_if(sample_size, -1)),

    # "Not included in poll" is structural missingness, not 0%.
    Other = as.numeric(na_if(Other, "Not included in poll")),

    # Only recode NA to "None" where affiliation is genuinely not applicable.
    partisan_affiliation = case_when(
      partisanship == "Nonpartisan" & is.na(partisan_affiliation) ~ "None",
      TRUE ~ partisan_affiliation
    ),

    # poll_id can be repeated across states, so create a state-poll key.
    poll_key = str_c(year, poll_id, state, sep = "_")
  ) |>
  select(-contest)

polls_2016_clean <- polls_2016_raw |>
  extract(
    poll_info,
    into = c("year", "state", "contest"),
    regex = "^(\\d{4})-(.*?)-(president|presidential-general-election)-.*$",
    remove = FALSE
  ) |>
  mutate(
    year = as.integer(year),

    # Make Washington DC consistent with the 2012 poll file.
    state = str_replace(state, "^washington-d-c$", "washington-dc"),

    start_date = as.Date(start_date),
    end_date = as.Date(end_date),
    days_to_election = as.integer(election_day["2016"] - end_date),

    poll_id = as.character(poll_id),
    sample_size = as.integer(na_if(sample_size, -1)),
    Other = as.numeric(na_if(Other, "Not included in poll")),

    partisan_affiliation = case_when(
      partisanship == "Nonpartisan" & is.na(partisan_affiliation) ~ "None",
      TRUE ~ partisan_affiliation
    ),

    poll_key = str_c(year, poll_id, state, sep = "_")
  ) |>
  select(-contest)

glimpse(polls_2012_clean)
glimpse(polls_2016_clean)
```

```{r}
#| label: time-plot
#| fig-cap: "Temporal distribution of polling effort, 2016."
polls_2016_clean |>
  ggplot(aes(days_to_election)) +
  geom_histogram(binwidth = 7) +
  scale_x_reverse() +
  labs(x = "Days to election (binned by week)", y = "Number of polls")
```

```{r}
#| label: space-coverage
bind_rows(
  polls_2012_clean |> count(state, name = "n_polls") |> mutate(year = 2012),
  polls_2016_clean |> count(state, name = "n_polls") |> mutate(year = 2016)
) |>
  arrange(year, desc(n_polls))
```

::: {.callout-note}
## Discuss
Is polling effort uniform over time? Which states are heavily polled and which
are barely polled — is that random, or related to competitiveness? What does it
imply for any nationwide statement? Do your conclusions hold for 2012 as well?
:::

*Your answer:*

---

# Cleaning, tidying, and combining

::: {.callout-note}
## Discuss
The 2012 and 2016 poll files have **different candidate columns** (Obama/Romney
vs Trump/Clinton/Johnson/McMullin) and different column orders. To analyse them
together, reshape to a tidy long form: one row per (poll, candidate). Why is
long form easier to combine and model than the wide form supplied?
:::

- Easier to group/filter by candidate and year.
- Easier for graphs and statistical models.
- Candidates who did not run in a particular year can simply be absent rather than needing separate columns.

```{r}
#| label: reshape-polls

polls_2012_long <- polls_2012_clean |>
  pivot_longer(
    cols = all_of(candidate_cols_2012),
    names_to = "candidate",
    values_to = "pct"
  ) |>
  mutate(
    party_group = case_when(
      candidate == "Obama" ~ "Democrat",
      candidate == "Romney" ~ "Republican",
      candidate == "Undecided" ~ "Undecided",
      TRUE ~ "Other"
    )
  )

polls_2016_long <- polls_2016_clean |>
  pivot_longer(
    cols = all_of(candidate_cols_2016),
    names_to = "candidate",
    values_to = "pct"
  ) |>
  mutate(
    party_group = case_when(
      candidate == "Clinton" ~ "Democrat",
      candidate == "Trump" ~ "Republican",
      candidate == "Undecided" ~ "Undecided",
      TRUE ~ "Other"
    )
  )

glimpse(polls_2012_long)
glimpse(polls_2016_long)
```

Before combining, validate the cleaned files. Do not remove a row simply because
a `poll_id` is repeated: the same poll can contain results for multiple states.

```{r}
#| label: validate-before-bind

# 1. Did state extraction fail anywhere?
tibble(
  dataset = c("2012", "2016"),
  failed_state_extraction = c(
    sum(is.na(polls_2012_clean$state)),
    sum(is.na(polls_2016_clean$state))
  )
)

# 2. Is each state-poll key unique in the WIDE data?
bind_rows(
  polls_2012_clean |> count(poll_key) |> filter(n > 1) |> mutate(year = 2012),
  polls_2016_clean |> count(poll_key) |> filter(n > 1) |> mutate(year = 2016)
)

# 3. Are any reported candidate percentages outside the possible 0-100 range?
bind_rows(
  polls_2012_long |>
    filter(!is.na(pct), !between(pct, 0, 100)) |>
    mutate(source_year = 2012),
  polls_2016_long |>
    filter(!is.na(pct), !between(pct, 0, 100)) |>
    mutate(source_year = 2016)
)

# 4. Flag polls whose reported response categories sum well above 100%.
# Allow a small tolerance for rounding; investigate rather than auto-delete.
bind_rows(
  polls_2012_clean |>
    mutate(
      reported_total = rowSums(
        across(all_of(candidate_cols_2012)),
        na.rm = TRUE
      )
    ) |>
    filter(reported_total > 102) |>
    select(year, poll_key, poll_info, reported_total),
  polls_2016_clean |>
    mutate(
      reported_total = rowSums(
        across(all_of(candidate_cols_2016)),
        na.rm = TRUE
      )
    ) |>
    filter(reported_total > 102) |>
    select(year, poll_key, poll_info, reported_total)
)
```

Now combine the datasets. Keep both versions:

- `polls_clean`: one row per state-poll, for polling effort/coverage analyses.
- `polls_long`: one row per state-poll-candidate, for candidate/support analyses.

```{r}
#| label: combine-polls

polls_clean <- bind_rows(
  polls_2012_clean,
  polls_2016_clean
)

polls_long <- bind_rows(
  polls_2012_long,
  polls_2016_long
)

glimpse(polls_clean)
glimpse(polls_long)
```

---

# Missingness — what is missing, and how is it recorded?

Analyse the missingness in the combined polling dataset.

::: {.callout-note}
## Discuss
- Is missing `Undecided` a true zero (nobody undecided) or an unknown (not
  reported)? How would you check, and why does it matter?
- A negative `sample_size` is impossible — what value is being used as a
  sentinel for "unknown"? What happens to your means/weights if you leave it in?
- What non-numeric entries appear in `Other`, and what will you do with them?
- For each affected variable: is it MCAR, MAR, or MNAR as far as you can tell,
  and will you drop, recode, impute, or model around it? Every fix lives in
  code, not in the raw file.
:::

Undecided
- Missing in:
+ 2012: 132 rows
+ 2016: 128 rows
- More likely it was not reported or not included.
- Likely MAR / related to poll design
- Keep as NA; do not replace with zero automatically.

sample_size
- Both 2012 and 2016 contain one -1 value.
- -1 is a sentinel for an unknown sample size. Recode -1 → NA.
- Leaving it as -1 would distort means and especially weighting calculations

Other
- "Not included in poll" occurs:
+ 2012: 567 times
+ 2016: 507 times
- Recode this to NA, not zero.
- This is mostly structural / poll-design missingness.

Johnson and McMullin
- Johnson missing: 975/1,303 rows.
- McMullin missing: 1,286/1,303 rows.
- Usually means the candidate was not included in that poll, not that support was zero.
- Treat as structural missingness.

partisan_affiliation
- Missing for nonpartisan polls.
- This is not really missing information; it means “not applicable”.
- Could recode to "None" for analysis.
```{r}
# Check that nonpartisan polls are represented explicitly as "None".
polls_clean |>
  count(year, partisanship, partisan_affiliation, sort = TRUE)
```


General rule
- Do not replace all NAs with zero.
- First identify why the value is missing.

```{r}
#| label: missingness-audit

# Missing candidate percentages by year and candidate.
polls_long |>
  group_by(year, candidate) |>
  summarise(
    n_rows = n(),
    n_missing = sum(is.na(pct)),
    pct_missing = 100 * mean(is.na(pct)),
    .groups = "drop"
  ) |>
  arrange(year, desc(pct_missing))

# Check missingness in the main poll-level fields used later.
polls_clean |>
  summarise(
    across(
      c(state, start_date, end_date, sample_size,
        mode, sample_subpopulation, partisanship, partisan_affiliation),
      ~ sum(is.na(.x))
    )
  )
```

---

# Bias — who/what is over- or under-represented?

```{r}
#| label: bias-pollster
polls_2016_raw |> count(pollster, sort = TRUE) |> head(15)
```


::: {.callout-note}
## Discuss
- **Coverage:** are the polls a representative sample of *states*, or
  concentrated on a few?
- **Pollster concentration:** if a few firms dominate the rows, whose house
  methods are over-weighted, and how might you account for it (grouping,
  weighting, a pollster effect)?
- **Partisanship:** do partisan-sponsored polls differ systematically? Nuisance
  to remove, or variable to study?
- **Selection:** the data is what someone *chose* to collect (which states,
  which weeks). Name one conclusion this selection could distort.
:::

Coverage bias
- States are not equally represented.
- Battleground states have many more polls.

Pollster concentration
- A few pollsters contribute many observations.
- Example:
+ 2012: Rasmussen 138, PPP 128.
+ 2016: Ipsos/Reuters 167, SurveyMonkey 152.
- Their particular methods could therefore have too much influence.
- We could include a pollster effect, group by pollster, or consider weighting.

Partisan bias
- Some polls are sponsored or conducted by partisan organisations.
- We should test whether these polls systematically favour one candidate.
- partisanship can therefore be an explanatory variable, not simply something to delete.

Selection bias
- Pollsters choose which states and when to poll.
- For example, a simple average of all rows could make national opinion look more like opinion in heavily polled swing states.

Overall
- The polling dataset is not a random sample of states, dates, or pollsters.
- So a simple average across all rows could be misleading.

---

# Wrangling data

This `.qmd` file can serve as a place to experiment with data cleaning and exploration. You can add/change it in future forums and when working on the EDA Report assessment.

For clarity and reproducibility you should create a master `poll-cleaning.R` script which starts with the raw data files and creates the cleaned, combined versions of the datasets you will perform EDA on. The code in this file serves as a starting point for this master cleaning script. 

The basic structure of the master cleaning script should be as follows:

```{r}
#| label: read-clean-combine-write
#| eval: false

# load packages
library(...)

# set up directories
raw_data_dir <- here(...)
clean_data_dir <- here(...)

# read in raw data
raw_data <- read_csv(raw_data_dir...)

# clean datasets
clean_data <- raw_data |> mutate(...)

# combine datasets as required
combinded_data <- bind_rows(...)
# save datasets
write_csv(clean_data_dir...)
```

All subsequent EDA and analysis should use the cleaned data.
This document serves as a record of the cleaning and EDA you have tried (but not necessarily included in your report).